---
title: "Converting from wide to long format & ANOVA in R"
subtitle: "Lab Code Challenge 3"
author: "Toka"
date: 2018-09-25T21:13:14-05:00
categories: ["R"]
tags: ["R Markdown", "plot", "lab code challenge"]
---

##Introduction

In this post, I will discuss about how to convert data from wide format to long format; how to create a 2D figure with information from 3 variables; and how to conduct ANOVA in 3 different ways.

The data in today's example was collected by our lab (led by Dr. Paul Minda) for the purpose of predicting category-learning on two types of categories (i.e.,RD: Rule-defined, and II: Information Integration). Columns from B to AY are all predictive variables. Columns from AZ to BC are the % correct (i.e., performance) in the 4 blocks of trails from either RD or II. Participants were assigned in either RD or II for all 4 blocks. Column BD is the average performance across the 4 blocks for each participant. We will be taking specific predictive variables as examples to explore their relationship with the outcome variable (i.e., performance)

<b>Below are the specific instructions we will be following:</b>

* First, create a new data frame that takes the column with performance in each block (AZ-BC) and puts them into a single column called PERFORMANCE and the a new column called “BLOCK” with 1,2,3,4 as values. This new frame will be four times taller, but does not need to include every variables, only the ones you need.

* Generate a line plot with errorbars (or points) that show PERFORMANCE with BLOCK as X-Axis and different lines for Category. Decide on the best way to also show MONTH (same plot different line, or separate plots)

* Run a mixed factor ANOVA on PERFORMANCE with BLOCK as within and CATEGORY and MONTH and between. You can do the same for CATEGORY and TIME, and even try the full model: PERFORMANCE with BLOCK (within) and CATEGORY, MONTH, and TIME and between.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r echo=FALSE}
setwd('/Users/tzhu9/Documents/OneDrive - The University of Western Ontario/Blogdown/tokaalmighty91.github.io/content/post')
```

##Preparation
###Install Packages
```{r message=FALSE, warning=FALSE, results='hide'}
#install.packages('prettydoc')
#install.packages('lme4')
#install.packages('ez', dependencies = TRUE)
#install.packages('data.table', dependencies= TRUE)
#install.packages('afex')
```
###Load packages
```{r message=FALSE, warning=FALSE, results='hide'}
library(prettydoc)
library(rmarkdown)
library(Rmisc)
library(readxl)
library(ggplot2)
library(data.table)
library(ez)
library(knitr)
library(tidyr)
library(tidyverse)
library(lme4)

```

##Data Manipulation
###Subsetting and Relabelling

Instead of dragging hundreds predictive variables), let's just pick a few to play with. I selected 'SubjNum', 'Cat', 'Month', 'DayofWeek', 'Time2' as predictive variables in our new dataframe, and of course, we need performance (i.e., '1_Block', '2_Block', '3_Block', '4_Block', 'Total') as our outcome variables.

I renamed some columns so the header row of dataframe is not cramped up together.
```{r}
data<-read_excel('ModifiedFullData.xlsx')

# Select relevant columns
data<-data[c('SubjNum','Cat','Month','DayofWeek','Time2',
            '1_Block','2_Block','3_Block','4_Block','Total')]

# Rename columns
data<-rename(data,
            Ss=SubjNum,
            Cat=Cat,
            Mon=Month,
            W_day=DayofWeek,
            Time=Time2,
            '1'='1_Block',
            '2'='2_Block',
            '3'='3_Block',
            '4'='4_Block',
            'Total'='Total')

# It's important to convert Subject into factor
# So R doesn't treat the entries as numerics
data$Ss <- factor(data$Ss)

# Drop May data due to incomplete comparison groups
data <- subset(data, Mon != "05_May")

# View head of data
kable(head(data))
```

###Changing Data from Wide to Long Form

Here we want to create a new data frame that takes the column with performance in each block (AZ-BC) and puts them into a single column called 'accuracy' and the a new column called “block” with 1,2,3,4 as values. This new frame will be four times taller, but does not need to include every variables, only the ones you need.
```{r}
# Transform data from wide to long form using 'tidyr' package
data_long <- gather(data, block, accuracy, '1':'4', 
                    factor_key=TRUE)

# Make sure block is a factor
data_long$block=factor(data_long$block)
kable(head(data_long))
```

##Data Visualization

###Plotting performance change across blocks

Generate a line plot with errorbars (or points) that show 'accuracy' with 'block' as X-Axis and different lines for Category. 

First, let's create a summary table which we would base our graph on.
```{r}
# Create summary table for block performance on each category type
summary_Block<-summarySE(data_long,                               
                         measurevar='accuracy',
                         groupvars=c('Cat','block'))

kable(summary_Block)
```

```{r}
# Plot accuracy change across blocks
ggplot(summary_Block, aes(block, accuracy, 
            color= Cat, group= Cat)) + 
    geom_line() +  
  
# add errorbar
    geom_errorbar(aes(ymin=accuracy-se, ymax=accuracy+se),
                colour="black", width=.1) +
  
# add labels and titles
    labs(x= "Block", y= "Accuracy") +
    scale_colour_hue(name="Category Type",   
                     breaks=c("RD", "II"),
                     labels=c("Rule_Defined", 
                              "Information_Integration")) +                   
    ggtitle("Performance Change across Blocks") +
    theme(plot.title= element_text(hjust= 0.5)) +
  
# set range of y
    ylim(.5, .85) +                        
    theme_light() +
  
# Position legend in bottom right
    theme(legend.justification=c(1,0),
          legend.position=c(1,0))               
```

###Plotting 4 Blocks' Performance change across the Months

Now, we also want to display MONTH information on our Category-Accuracy figure. First, let's create a summary table with Month information included.
```{r}
# Create summary table for plotting accuracy change across months
summary_Month<-summarySE(data_long,                               
                         measurevar='accuracy',
                         groupvars=c('block','Mon','Cat'))

kable(head(summary_Month))
```

There're many ways to go about creating 2D figure that shows information from 3 variables. I chose to create 2 side-by-side figures seperated by category-type, with Month as x-axis, and accuracy as y-axis, the 4 blocks' performance are represented in different colors.  
```{r}
# Plot individual block performance across the months
# Figures are seperated by Category Type
ggplot(summary_Month, aes(Mon, accuracy, 
            color= block)) +
    geom_line(aes(group = block)) +
    geom_point(aes(shape = block), size = 3) +
    facet_wrap(~Cat) +
    scale_x_discrete(labels = c("01_Jan" = "Jan", "02_Feb" = "Feb",
                              "03_Mar" = "Mar", "04_Apr" = "Apr",
                              "05_May" = "May", "09_Sep" = "Sept",
                              "10_Oct" = "Oct", "11_Nov" = "Nov"))+
  
    theme_update(plot.title= element_text(hjust= 0.8)) +
    ggtitle("Block Performance across Different Months") +
  
# set range of y
    ylim(.5, .85) +                        
    theme_light() 
```

####Create a summary table of data

I'm creating a summary table to see N for each month because the low RD performance in November looks suspicious to me. We should never interpret a graph without knowing the sample size. Upon inspection of the summary table, we can see that only 4 Ss participated in Nov_RD condition, we now understand this low performance should not be trusted due to the tiny sample size.
```{r}
summary_data<-summarySE(data,                               
                         measurevar='Total',
                         groupvars=c('Mon','Cat'))

kable(summary_data)

# From this table we can see there were only 4 participants assigned to
# RD category in Nov, therefore, it is too small of a sample size to
# say that people perform worse in Nov.
```

##Mixed ANOVAs 

Run a mixed factor ANOVA on 'accuracy' with 'block' as within and CATEGORY and MONTH and between. You can do the same for CATEGORY and TIME, and even try the full model: 'accuracy' with 'block' (within) and CATEGORY, MONTH, and TIME and between.

###Using aov function
```{r}
# mixed ANOVA using aov: with 'block' as within-IV, 'Cat','Month','Time' as between-IV, accuracy is DV
anova1<-aov(accuracy~block*Cat*Mon*Time+
            Error(Ss/block)+(Cat*Mon*Time),
            data=data_long)
summary(anova1)
```

###Using ezANOVA. Returns 1 or more cell with missing data error.
```{r}
# ezANOVA returns error
# anova2<-ezANOVA(data= data_long,
#                 dv= .(accuracy),
#                 wid= .(Ss),
#                 within= .(block),
#                 between= .(Month, Time, Cat),
#                 type= 3, detailed= TRUE)
```

###Using lme4.
```{r message=FALSE, warning=FALSE}
# 'block', 'Cat', 'Month', 'Time' are fixed effects
# 'Ss' is random effects
fullmodel_lmer <- lmer(accuracy ~ block + Cat + Mon + Time +
                       (1|Ss), data_long, REML=FALSE)

# Check how to interpret summary output
summary(fullmodel_lmer)
```











